Negative Selection in the 20th Century

### 2020-05-14 why-selection-against-EA.Rmd
Initial commit

knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = draft, 
        error = draft
      )
theme_set(theme_minimal())

order_abs <- function (x) x[1] # for use when reordering factors

@@

@@

@@

```{r run-regressions}
# for each score, calculate the correlation with number of siblings
# 
run_regs <- function (dep_var, score_name) {
  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
  reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE)

@@

```{r}
run_regs_time <- function(children, score_name, period) {
  dep_var  <- if (children) "log(n_children + 1)" else "log(n_sibs + 2)"
  year_var <- if (children) "year_flb" else "YOB"
  famhist$in_period <- ! is.na(famhist[[year_var]]) & famhist[[year_var]] %in% period:(period+4)

@@

pars <- expand_grid(
        children = c(TRUE, FALSE),
        score_name = score_names, 
        period = seq(1940, 2000, 5)
      )
res_yr <- pmap_dfr(pars, run_regs_time)

@@

```{r over-time}

ggplot(res_yr, aes(period, estimate, colour = children)) + 
      geom_linerange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
      geom_line() + 
      facet_wrap(vars(term))
```

Selection over time appears to be decreasing, if we focus on raw PGS. Figure
\@ref(fig:effect-sizes-by-period) shows effect sizes for  survey respondents
born in the 1930s and 1940s, versus respondents born in the 1950s to 1970s.

```{r effect-sizes-by-period, fig.cap = "Effect sizes of PGS on n children by period. Significance tests compare early to late effect sizes", fig.width = 6, fig.height = 7}

famhist$period <- chop(famhist$YOB, c(1930, 1950, 1974), 
      labels = c("30s_40s", "50s_70s"))

run_time_regs <- function(score_name) {

  fml <- as.formula(glue("log(n_sibs+1) ~ {score_name}:period"))
  mod <- lm(fml, famhist)
  res <- tidy(mod)
  res <- filter(res, ! term %in% c("(Intercept)"))
  res$term <- gsub(".*:period", "", res$term)

  glhts <- list(
    glht(mod, glue("{score_name}:period30s_40s - {score_name}:period50s_70s = 0"))
  )
  glhts <- glhts %>% map(summary) %>% map(tidy) %>% map_dbl("p.value")
  res$glht.p.value <- c(NA, glhts)

  res
}

names(score_names) <- score_names
time_regs <- map_dfr(score_names, run_time_regs, .id = "reg")

time_regs %<>% 
      rename(period = term) %>% 
      mutate(reg = fct_reorder(reg, estimate, first))

time_regs_wide <- time_regs %>% 
      select(reg, period, estimate, glht.p.value) %>% 
      pivot_wider(reg, names_from = period, 
        values_from = c(estimate, glht.p.value)) %>% 
      mutate(Significant = glht.p.value_50s_70s < 0.05/(nrow(time_regs) * 3/4))

time_regs %>% 
      group_by(reg) %>% 
      mutate(
        Significant = na.omit(glht.p.value) < 0.05/(nrow(time_regs) * 3/4)
      ) %>% 
      ggplot(aes(x = reg)) +
        geom_hline(yintercept = 0, colour = "darkgrey") +
        geom_linerange(aes(
          ymin = estimate_30s_40s, 
          ymax = estimate_50s_70s, colour = Significant), 
          data = time_regs_wide) + 
        geom_point(aes(y = estimate, fill = period, colour = Significant), 
          shape = 21, size = 2) + 
        scale_colour_manual(values = c("grey", "orange3")) +
        scale_fill_brewer(type = "qual", palette = 3) +
        coord_flip()
```

# Causality

To be selected against, an allele does not need to cause fewer children, only to
be correlated with fewer children. Nevertheless, it is interesting to ask
whether individual PGS caused sample members' parents to have fewer children. As
a first cut, we regress log(number of siblings _+ 1) on all PGS together. We run
two regressions: one controlling for genetic principal components, and one
without these controls. Figure \@ref(fig:multivariate-reg-plots) shows the results.

```{r multivariate-regs}

# EA3 and EA2 have a 0.9 raw correlation. These are the only pair of PGS with
# a raw correlation of more than 0.8
reg_score_names <- setdiff(score_names, "EA2_noUKB")

without_pcs <- paste0("log(n_sibs+1) ~ ", paste(reg_score_names, collapse = " + "))
fml <- as.formula(without_pcs)
fml_with_pcs <- as.formula(paste(without_pcs, "+", paste0("PC", 1:40, 
      collapse = " + ")))

all_mod <- lm(fml, famhist)
all_res <- tidy(all_mod, conf.int = TRUE)

all_mod_pcs <- lm(fml_with_pcs, famhist)
all_res_pcs <- tidy(all_mod_pcs, conf.int = TRUE)

all_res <- bind_rows(no = all_res, yes = all_res_pcs, .id = "PCs")
all_res <- filter(all_res, term != "(Intercept)", ! grepl("^PC", term))

```{r multivariate-reg-plots, fig.cap = "Multivariate regressions. Significance tests compare effect sizes to zero."}

all_res$term <- fct_reorder(all_res$term, all_res$estimate, order_abs)

n_regs <- as.double(nrow(all_res)) # avoids an ugly "L"
ggplot(all_res, aes(term)) + 
      geom_point(aes(y = estimate, fill = PCs, 
        colour = p.value <= 0.05/{{n_regs}}), shape = 21) + 
      scale_colour_manual(values = c("grey", "black")) +
      geom_hline(yintercept = 0, colour = "darkgrey") +
      labs(x = "PGS", y = "Effect size") +
      coord_flip() 
```

>>>>

options(digits = 2)
knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = draft, 
        error = draft
      )
theme_set(theme_minimal())

order_abs <- function (x) x[1] # for use when reordering factors
pretty_names <- function (name) {
  name <- gsub("_", " ", name)
  name <- strsplit(name, " ")
  map_chr(name, 
        ~paste0(
          toupper(substring(.x, 1, 1)), 
          substring(.x, 2),
          collapse = " "
        ))
}

@@

reverse_code <- c("agreeableness", "age_at_menarche", "age_at_menopauze",
  "cognitive_ability", "conscientiousness", "EA2_noUKB",
  "EA3_excl_23andMe_UK", "extraversion", "height_combined",
  "openness")
famhist[reverse_code] <- famhist[reverse_code] * -1

@@

# Data

Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
variance 1. Scores for educational attainment, cognitive ability, height,
extraversion, agreeableness, conscientiousness, age at menarche, and age at
menopause were reverse coded. High scores are therefore "negative" in the loose
sense that their target phenotypes would be considered undesirable by many
people.

@@

```{r run-regs}
# for each score, calculate the correlation with number of siblings
# 
run_regs <- function (dep_var, score_name) {
  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
  reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE)

@@

```{r run-regs-period}
run_regs_period <- function(children, score_name, period) {
  dep_var  <- if (children) "log(n_children)" else "log(n_sibs + 1)"
  year_var <- if (children) "year_flb" else "YOB"
  famhist$in_period <- ! is.na(famhist[[year_var]]) & 
                       famhist[[year_var]] %in% period:(period+9)

@@

pars <- expand_grid(
        children = c(TRUE, FALSE),
        score_name = score_names, 
        period = seq(1940, 2000, 10)
      )
res_yr <- pmap_dfr(pars, run_regs_period)

@@

```{r plot-by-period, fig.cap = "Effect sizes of PGS on number of children/siblings by ten year periods", fig.height = 7, fig.width = 6}
res_yr %>% 
      mutate(
        dep.var = ifelse(children, "Children", "Siblings")
      ) %>% 
ggplot(aes(period, estimate, group = dep.var)) + 
      geom_hline(yintercept = 0, colour = "grey50", size = 0.5) +
      geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.3, fill = "lightblue") +
      geom_line(aes(linetype = dep.var), colour = "navy") + 
      facet_wrap(
        vars(term), 
        scales = "free_y", 
        ncol = 4, 
        strip.position = "bottom"
      ) +
      theme(
        panel.grid.minor.y = element_blank(), 
        legend.title = element_blank()
      )
```

Selection over time appears to be decreasing. Figure \@ref(fig:plot-by-period) shows
effect sizes estimated within 10 year periods. Sibling regressions use the
respondent's own birth year. Children regressions use the year of the
respondent's first live birth. Since this is not available for men, only female
respondents are included. By definition, the sibling regressions exclude members
of the parents' generation who had no children. To make the effect size
estimates as comparable as possible, children regressions include only
respondents with at least one child ever born; the dependent variable is
*log(number of children)* or *log(total number of siblings)*.

Despite this, effect sizes tend to be smaller for children regressions. This
could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
if respondents themselves are a more selected sample than the respondents' 
parents.

## Subgroups

We next examine the correlates of natural selection by running regressions on
*log(number of children + 1)* within various subgroups of the sample.

```{r run-regs-subset}

run_regs_subset <- function(score_name, subset) {
  fml <- as.formula(glue("log(n_children + 1) ~ {score_name}"))
  mod <- lm(fml, famhist, subset = eval(subset))
  res <- tidy(mod, conf.int = TRUE)
  res <- filter(res, term == {{score_name}})
  res$subset <- list(subset)
  res
}

subsets <- rlang::exprs(sex == 0, sex == 1)
par <- expand_grid(score_name = score_names, subset = subsets)
res_subsets <- pmap_dfr(par, run_regs_subset)
```

```{r fn-plot-subsets}
plot_subsets <- function(ss_name, ss_expr, ...) {
  ss_name <- rlang::enquo(ss_name)
  subset_expr <- rlang::enexprs(...)
  res_subsets %>% 
        filter(subset %in% !!subset_expr) %>% 
        mutate(
          subset = as.character(subset), 
          !!ss_name := {{ss_expr}},
          term = fct_reorder(term, estimate, order_abs)
        ) %>% 
        ungroup() %>% 
        ggplot(aes(term, estimate, group = !!ss_name, colour = !!ss_name)) +
          geom_hline(yintercept = 0, colour = "grey20") +
          geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.7) +
          coord_flip() +
          scale_colour_brewer(palette = 2, type = "qual")
}
```

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. Education and height PGS are strongly
negatively selected among women.

```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}
plot_subsets(
        Sex, 
        recode(subset, "sex == 0" = "Female", "sex == 1" = "Male"),
        sex == 0, sex == 1
      )
```

```{r plot-sd-years, eval = FALSE}
# inequality over time. Nothing to see, almost everything is flat.
grid <- expand_grid(score_name = score_names, YOB = 1940:1965)  
grid$sd <- pmap_dbl(grid, ~sd(famhist[famhist$YOB==.y, ][[.x]], na.rm = TRUE))
ggplot(grid, aes(YOB, sd)) + 
      geom_line() + 
      facet_wrap(~score_name, scales = "free_y", ncol = 4)
```
